% -*- mode: latex; TeX-master: "Vorbis_I_spec"; -*-
%!TEX root = Vorbis_I_spec.tex
% $Id$
\section{comment field and header specification} \label{vorbis:spec:comment}

\subsection{Overview}

The Vorbis text comment header is the second (of three) header
packets that begin a Vorbis bitstream. It is meant for short text
comments, not arbitrary metadata; arbitrary metadata belongs in a
separate logical bitstream (usually an XML stream type) that provides
greater structure and machine parseability.

The comment field is meant to be used much like someone jotting a
quick note on the bottom of a CDR. It should be a little information to
remember the disc by and explain it to others; a short, to-the-point
text note that need not only be a couple words, but isn't going to be
more than a short paragraph.  The essentials, in other words, whatever
they turn out to be, eg:

\begin{quote}
Honest Bob and the Factory-to-Dealer-Incentives, \textit{``I'm Still
Around''}, opening for Moxy Fr\"{u}vous, 1997.
\end{quote}




\subsection{Comment encoding}

\subsubsection{Structure}

The comment header is logically a list of eight-bit-clean vectors; the
number of vectors is bounded to $2^{32}-1$ and the length of each vector
is limited to $2^{32}-1$ bytes. The vector length is encoded; the vector
contents themselves are not null terminated. In addition to the vector
list, there is a single vector for vendor name (also 8 bit clean,
length encoded in 32 bits). For example, the 1.0 release of libvorbis
set the vendor string to ``Xiph.Org libVorbis I 20020717''.

The vector lengths and number of vectors are stored lsb first, according
to the bit packing conventions of the vorbis codec. However, since data
in the comment header is octet-aligned, they can simply be read as
unaligned 32 bit little endian unsigned integers.

The comment header is decoded as follows:

\begin{programlisting}
  1) [vendor\_length] = read an unsigned integer of 32 bits
  2) [vendor\_string] = read a UTF-8 vector as [vendor\_length] octets
  3) [user\_comment\_list\_length] = read an unsigned integer of 32 bits
  4) iterate [user\_comment\_list\_length] times {
       5) [length] = read an unsigned integer of 32 bits
       6) this iteration's user comment = read a UTF-8 vector as [length] octets
     }
  7) [framing\_bit] = read a single bit as boolean
  8) if ( [framing\_bit] unset or end-of-packet ) then ERROR
  9) done.
\end{programlisting}




\subsubsection{Content vector format}

The comment vectors are structured similarly to a UNIX environment variable.
That is, comment fields consist of a field name and a corresponding value and
look like:

\begin{quote}
\begin{programlisting}
comment[0]="ARTIST=me";
comment[1]="TITLE=the sound of Vorbis";
\end{programlisting}
\end{quote}

The field name is case-insensitive and may consist of ASCII 0x20
through 0x7D, 0x3D ('=') excluded. ASCII 0x41 through 0x5A inclusive
(characters A-Z) is to be considered equivalent to ASCII 0x61 through
0x7A inclusive (characters a-z).


The field name is immediately followed by ASCII 0x3D ('=');
this equals sign is used to terminate the field name.


0x3D is followed by 8 bit clean UTF-8 encoded value of the
field contents to the end of the field.


\paragraph{Field names}

Below is a proposed, minimal list of standard field names with a
description of intended use.  No single or group of field names is
mandatory; a comment header may contain one, all or none of the names
in this list.

\begin{description} %[style=nextline]
\item[TITLE]
	Track/Work name

\item[VERSION]
	The version field may be used to differentiate multiple
versions of the same track title in a single collection. (e.g. remix
info)

\item[ALBUM]
	The collection name to which this track belongs

\item[TRACKNUMBER]
	The track number of this piece if part of a specific larger collection or album

\item[ARTIST]
	The artist generally considered responsible for the work. In popular music this is usually the performing band or singer. For classical music it would be the composer. For an audio book it would be the author of the original text.

\item[PERFORMER]
	The artist(s) who performed the work. In classical music this would be the conductor, orchestra, soloists. In an audio book it would be the actor who did the reading. In popular music this is typically the same as the ARTIST and is omitted.

\item[COPYRIGHT]
	Copyright attribution, e.g., '2001 Nobody's Band' or '1999 Jack Moffitt'

\item[LICENSE]
	License information, eg, 'All Rights Reserved', 'Any
Use Permitted', a URL to a license such as a Creative Commons license
("www.creativecommons.org/blahblah/license.html") or the EFF Open
Audio License ('distributed under the terms of the Open Audio
License. see http://www.eff.org/IP/Open\_licenses/eff\_oal.html for
details'), etc.

\item[ORGANIZATION]
	Name of the organization producing the track (i.e.
the 'record label')

\item[DESCRIPTION]
	A short text description of the contents

\item[GENRE]
	A short text indication of music genre

\item[DATE]
	Date the track was recorded

\item[LOCATION]
	Location where track was recorded

\item[CONTACT]
	Contact information for the creators or distributors of the track. This could be a URL, an email address, the physical address of the producing label.

\item[ISRC]
	International Standard Recording Code for the
track; see \href{http://www.ifpi.org/isrc/}{the ISRC
intro page} for more information on ISRC numbers.

\end{description}



\paragraph{Implications}

Field names should not be 'internationalized'; this is a
concession to simplicity not an attempt to exclude the majority of
the world that doesn't speak English. Field \emph{contents},
however, use the UTF-8 character encoding to allow easy representation
of any language.

We have the length of the entirety of the field and restrictions on
the field name so that the field name is bounded in a known way. Thus
we also have the length of the field contents.

Individual 'vendors' may use non-standard field names within
reason. The proper use of comment fields should be clear through
context at this point.  Abuse will be discouraged.

There is no vendor-specific prefix to 'nonstandard' field names.
Vendors should make some effort to avoid arbitrarily polluting the
common namespace. We will generally collect the more useful tags
here to help with standardization.

Field names are not required to be unique (occur once) within a
comment header.  As an example, assume a track was recorded by three
well know artists; the following is permissible, and encouraged:

\begin{quote}
\begin{programlisting}
ARTIST=Dizzy Gillespie
ARTIST=Sonny Rollins
ARTIST=Sonny Stitt
\end{programlisting}
\end{quote}







\subsubsection{Encoding}

The comment header comprises the entirety of the second bitstream
header packet.  Unlike the first bitstream header packet, it is not
generally the only packet on the second page and may not be restricted
to within the second bitstream page.  The length of the comment header
packet is (practically) unbounded.  The comment header packet is not
optional; it must be present in the bitstream even if it is
effectively empty.

The comment header is encoded as follows (as per Ogg's standard
bitstream mapping which renders least-significant-bit of the word to be
coded into the least significant available bit of the current
bitstream octet first):

\begin{enumerate}
 \item
  Vendor string length (32 bit unsigned quantity specifying number of octets)

 \item
  Vendor string ([vendor string length] octets coded from beginning of string to end of string, not null terminated)

 \item
  Number of comment fields (32 bit unsigned quantity specifying number of fields)

 \item
  Comment field 0 length (if [Number of comment fields] $>0$; 32 bit unsigned quantity specifying number of octets)

 \item
  Comment field 0 ([Comment field 0 length] octets coded from beginning of string to end of string, not null terminated)

 \item
  Comment field 1 length (if [Number of comment fields] $>1$...)...

\end{enumerate}


This is actually somewhat easier to describe in code; implementation of the above can be found in \filename{vorbis/lib/info.c}, \function{\_vorbis\_pack\_comment()} and \function{\_vorbis\_unpack\_comment()}.






